---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Let's start the analysis by loading the dataset that is provided with the question. The dataset is provided as a compressed comma-separated values file from [this location](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). To start the analysis, we'll download the file, and uncompress it to the `dataset` folder.

```{r echo=TRUE}
library(dplyr)
library(ggplot2)
```

```{r echo=TRUE}
# Check that the files exist
datasetFolder <- "dataset"
datasetFile <- paste(datasetFolder, "activity.csv", sep = "/")
if (!file.exists(datasetFolder) || !file.exists(datasetFile)) {
    warn("Missing dataset file, this analysis will fail.")
}

# Load the data into memory
data <- read.csv(datasetFile, header = TRUE)
dataWithoutNA <- data[!is.na(data$steps),]

# Aggregate the number of steps per day
stepsPerDay <- dataWithoutNA %>%
  group_by(date) %>%
  summarize(totalSteps = sum(steps))
```

## What is mean total number of steps taken per day?

Now that we have the data loaded, we can start answering the questions. First of all, let's look at the distribution of the number of steps, by plotting a histogram of the number of steps per day. In this chart let's also add a line indicating the mean of the total steps.

```{r echo=TRUE}
g <- ggplot(stepsPerDay, aes(x = totalSteps))
g <- g + geom_histogram(binwidth = 1000, fill = "white", color = "black", size = 1)
g <- g + geom_vline(color = "red", aes(xintercept = mean(totalSteps)))
g
```

The chart shows the mean a little under 11000 steps, let's get the accurate value:

```{r echo=TRUE}
mean(stepsPerDay$totalSteps)
median(stepsPerDay$totalSteps)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
